---
title: "Assignment 9: Mapping"
author: "Jack Carpenter"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on mapping

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/NDWEUu73LooFJPVM8 

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-04-12

## Setup

1. Load the tidyverse, LAGOSNE, maps, dataRetrieval, sf, and nhdplusTools packages. Set your ggplot theme (can be theme_classic or something else).


```{r setup}
#check wd
getwd()

#load packages
library(tidyverse)
library(LAGOSNE)
library(maps)
library(dataRetrieval)
library(sf)
library(nhdplusTools)

#set ggplot theme
theme_set(theme_classic()+
            theme(axis.text = element_text(color = "black", size = 10),
                  legend.position = "right"))
options(scipen = 4)

```

## LAGOS-NE

2. Choose five lakes in the LAGOS-NE database that are located within the same state and have chlorophyll data. Subset your data accordingly, and create two plots:

* A map of the five lakes within the state boundary, with each lake point as its own color.
* A boxplot with jittered points layered on top of chlorophyll concentrations in each lake (chlorophyll on y axis, lake on x axis), with each lake having a fill and/or color that matches the map. 

```{r}
#first load LAGOS data
LAGOSdata <- lagosne_load()

#load LAGOS dataframes
LAGOSlocus <- LAGOSdata$locus
LAGOSstate <- LAGOSdata$state
LAGOSnutrient <- LAGOSdata$epi_nutr

#joined at the hip
LAGOScombined <- 
  left_join(LAGOSnutrient, LAGOSlocus) %>%
  left_join(., LAGOSstate) %>%
  select(lagoslakeid, sampledate, chla, nhd_lat, nhd_long, state)#choose data


#state your name
states <- st_as_sf(map(database = "state", plot = TRUE, fill = TRUE, 
                       col = "white"))
states.NH <- filter(states, ID == "new hampshire")

ggplot(states.NH) +
  geom_sf(fill = "white") #double check that it'll map

LAGOS.NH <- LAGOScombined %>%
  group_by(lagoslakeid) %>%
  filter(state == "NH") %>%
  drop_na(chla)
  
#choose 5 random lakes
samplelakes <- sample(x = LAGOS.NH$lagoslakeid, size = 5)

NH.Sample <- LAGOS.NH %>%
  filter(lagoslakeid %in% samplelakes)


#map it up, fuzzball
NH.sample.spatial <- st_as_sf(NH.Sample, coords= c("nhd_long", "nhd_lat"), 
                              crs = 4326)

ggplot() +
  geom_sf(data = states.NH, fill = "white") +
  geom_sf(data = NH.sample.spatial, aes(color = as.factor(lagoslakeid)), 
          alpha = 0.8, size = 1)+
  scale_color_viridis_d() +
  labs(color = "Lake ID")

ggplot(NH.Sample, aes(x = as.factor(lagoslakeid), y = chla,
                   fill = as.factor(lagoslakeid))) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(alpha = 0.5) +
  scale_y_log10() +
  scale_color_viridis_d() +
  labs(x = "Lake ID", y = expression("Chlorophyll a ("*mu*"g/L)", fill = ""))+
  theme(legend.position = "none")

#I can't get the colors to match, even though I'm using the same color scale

```


## NHDPlus

3. Delineate and map the watershed and flowpaths in the network upstream of the Eno River gage in Durham (USGS-02085070). Your map should include all stream orders.


```{r}
#what do we have?
ENOparams <- whatNWISdata(siteNumbers = "02085070")

#where is it
ENOcoords <- ENOparams %>%
  select(site_no, dec_lat_va, dec_long_va) %>%
  distinct()

#define a starting point
start_point <- st_sfc(st_point(c(ENOcoords$dec_long_va, ENOcoords$dec_lat_va)),
                      crs = 4326)
start_comid <- discover_nhdplus_id(start_point)

#navigate the NLDI
NLDI <- navigate_nldi(list(featureSource = "comid", featureID = start_comid),
                      mode = "upstreamTributaries",
                      distance_km = 1000)

#watershed and flowpath - this is just the syntax that nhdplus needs
subset_file <- tempfile(fileext = ".gpkg") #needs the "." 
subset <- subset_nhdplus(comids = as.integer(NLDI$UT$nhdplus_comid),
                         output_file = subset_file,
                         nhdplus_data = "download",
                         flowline_only = FALSE,
                         return_data = TRUE,
                         overwrite = FALSE)

#create data frames from the subset list
flowlines <- subset$NHDFlowline_Network
catchment <- subset$CatchmentSP
# double check they each are sf, tbl, and df classes
#class(flowlines) 
#class(catchment)


#find gages near/in watershed
gages <- get_nwis(AOI = catchment) #AOI is area of interest
class(gages) #also as sf and df

gages <- st_intersection(gages, catchment) #now ONLY gages IN catchment

#now map it out
plot_nhdplus("USGS-02085070") #first way to map - easiest and broadest

ggplot() + #ggplot version, can control more aesthetics
  geom_sf(data = catchment, fill = "white", color = "grey", lwd = 0.5) +
  geom_sf(data = flowlines, aes(color = streamorde))+
  geom_sf(data = gages, color = "red", size = 1) +
  labs(color = "Stream Order")

#analysis
max(flowlines$totdasqkm) #total drainage area in km^2

summary(as.factor(flowlines$streamorde))

view(gages)
```

4. What is the upstream drainage area for this gage, and how are the flowpaths distributed with regard to stream order?

> The upstream drainage area is 367.68 square kilometers. Flowpaths are distributed with 123 first-order streams, 45 second-order streams, 19 thrid-order streams, and 28 fourth-order streams.

5. Are there any other gage sites in the upstream catchment? If so, where are they?
> Yes there are two more, in Hillsborough and at Cole Mill Rd near Huckleberry Spring.
